---
title: The Title of Your Study Here
subtitle: Add a Subtitle if Needed
author: Björn Lindberg
bibliography: bibliography.bib
csl: bmcemerg.csl
output: pdf_document
---


Introduction 
============
Trauma is defined as the clinical entity composed of both physical injury and the body’s associated response. [MeSH] Trauma patients are a heterogenous group with wide range of different causes where some of the most prominent ones are traffic related injuries, falls, self-harm, interpersonal violence and burns. It is a global health problem and the total number of injuries is estimated to be at least 520 million cases per year. [@James2020] The effects of trauma on patients and their families can be divided into non-fatal and fatal trauma.

The effects of non-fatal trauma can be subdivided into Disability Adjusted Life Years (DALY) lost and economical effects. DALY is a measurement using both the mortality in years lost between death and mean age and morbidity as in years lived with reduced function due to the disease. In 2017 trauma accounted for 13% of total DALY lost. [@GBD2017DALY] The economic effects in turn regards both the out of pocket expenses for treatments as well as reduced productivity due to reduced ability after the injury. A study in northern India from 2019 calculated that 22% of patients who presented with traumatic injuries afterward had 30% or more of their total expenditure on healthcare related costs and 12 % were pushed below the international poverty line as a consequence. [@Prinja2019]

Regarding the effect of traumatic injuries on mortality in 2017 trauma accounted for 8% of total global deaths. [@GBD2017DALY] According to calculations done by Mock et al. [@Mock2012] 90% of all deaths due to trauma were in low and middle income countries which also had a fatality rate 60-80% higher than that of high income countries and an improvement in trauma care globally to the level of that in high income countries could save about 1.7 lives annually, or about one third of the global deaths due to trauma.

It is known that most fatalities in response to traumatic injuries occur shortly after the injury and in order to reduce the mortality in patients with severe trauma it requires early intervention, often surgical, as by “the golden hour” concept. This is often due to patients suffering from internal and external bleeding where surgery is an important part of the hemorrhage control. [@King2019] It is evident that in parts of the world with fewer hospitals, reduced access to healthcare and a less developed ambulance networks the time parameter is one of the contributing factors to the difference in increased mortality as mentioned. 

For patients with severe infections the Sequential Organ Failure Assessment (SOFA) score was developed in 1994 to grade the level of organ dysfunction in critically ill patients. The SOFA score is based on the reduced function scored 0-4 for each of the respiratory, cardiovascular, neurological, hepatic, coagulation and renal systems. Initially the focus was mainly patients suffering from severe infections and sepsis, but it was early recognized to be applicable in a wide range of critical illnesses. [@Lambden2019] 

Alongside the last revision of the sepsis definition in 2016 the quick Sequential Organ Failure Assessment score (qSOFA score) was developed. The qSOFA score includes three parameters: Systolic blood pressure (SBP) below 100 mmHg, a respiratory rate (RR) above 22 min and an altered consciousness, measured with a Glasgow Coma Scale (GCS) score below 15. One point is awarded for each parameter yielding a total score of 0-3. Introduced as a screening tool, to be performed bedside to identify patients with high-risk of organ dysfunction due to infections. [@David2017],[@Shankar2016] It has lately also shown ability to predict mortality in patients with traumatic injuries when used both in prehospital care [@Miyamoto2019] as well as repeated evaluation during hospitalization [@Miyamoto2020].

Another tool to assess the severity of a patient’s condition due to trauma is the Revised Trauma Score (RTS) and the Triage Revised Trauma Score (T-RTS) composed of the same parameters as qSOFA, that is GCS, SBP and RR but with a score of 0-4 for each parameter instead of a single criteria. The scores can either be presented individually as used in RTS or added together to a total of 0-12 used in the T-RTS. The RTS thought of to be used in outcome evaluation and assessment of injury severity and the T-RTS for triage and finding patients a prehospital or initial setting in most need of specialized trauma care [@Champion1989] However there are some articles raising concerns about the sensitivity of the T-RTS depending on the type of injury. [@Giannakopoulos2011],[@Alvarez2016]

Because of the inherent urgency and complexity in trauma care it is vital to have an initial scoring system that is both simple to perform and that does not need laboratory nor other advanced measuring equipment. For this reason, the SOFA score can be said to use too advanced techniques and the T-RTS with its many subdividing cut-offs to be too complicated. The choice of using qSOFA is thus because of its promising signs in other studies and because of the possible clinical advantage in its simplicity.  

It should also be noted that the majority of the research done on the topic of trauma and trauma care has been performed in high income countries. It is thus not clear if the results are also applicable in low resource settings, which arguably stand to benefit the most of a scoring system to be able to prioritize the limited resources at hand. Furthermore, most studies report the predictive power of different scoring systems on mortality and it has not yet been sufficiently studied whether an initial qSOFA score can be used to predict admissions to an intensive care unit and thus aid in the direct clinical treatment of trauma patients.

<!-- just for the KI paper ---> 
## Healthcare in India 

The Indian healthcare is of now severely underfunded, the national healthcare expenditure is about 3.5% of national GDP, or about 70 USD per person in 2017, compared to a national healthcare expenditure of 5900 USD in Sweden for the same year. Out of these 70 USD per person 27 % is government expenditure and 72% is private expenditure, mostly out of pocket expenditures. (https://data.worldbank.org/) This makes the unregulated private healthcare sector almost 3 times as large as the public sector providing 80% of outpatient care and 60% of inpatient care in India. India also lacks healthcare personal and in many cases subspecialized medical training, especially regarding trauma care. (Nobhojit Roy)(Martin Gerdin)

Besides the deficient monetary aspects of Indias healthcare system it can also be said that there is a lack of sufficiently enforced safety policies and regulations, for example road traffic and workplace safety. Although it should be noted that in the new national health policy from 2017 one of the primary focuses are preventing rail and road traffic accidents. The national health policy also aims at expanding the number of hospital beds dedicated to trauma care across the country reducing the time between injury and care as guided by the golden hour concept discussed above. (Nobhojit Roy)(Martin Gerdin) 


## Aim  
The aim of this project is to assess the validity of qSOFA in predicting ICU admission in trauma patients admitted to hospitals in a low resource setting using data from the TITCO cohort [TITCO].


Methods
=======

## Source of data

<!-- 
4a) Describe the study design or source of data (e.g., randomized trial, cohort, or
registry data), separately for the development and validation data sets, if applicable
4b) Specify the key study dates, including start of accrual; end of accrual; and, if
applicable, end of follow-up. 

This study involved an analysis of the observational cohort Towards Improved Trauma Care Outcomes in India (TITCO), for which data were collected before the conception of this study. Ethics committees at all participating centres approved the collation of the database and granted a waiver of consent for patients with trauma (Lokmanya Tilak Municipal General Hospital, IEC/11/13; King Edward Memorial Hospital, IEC(I)/OUT/222/14; Seth Sukhlal Karnani Memorial Hospital, IEC/279; All‐India Institute of Medical Sciences, IEC/NP‐279/2013 RP‐01/2013). The study was conducted using anonymized data, and was registered at ClinicalTrials.gov (NCT03069755) before the research was undertaken.

-->

We conducted a retrospective analysis of the observational Towards Improved Trauma Care Outcomes in India (TITCO) cohort [@TITCO] <!-- TITCO collaborators (2017). TITCO dataset version 1. Available from https://github.com/titco/titco-I. (Behöver läggas in i .bib)-->. The data for TITCO was collected during july 2013 to december 2015 and contains patients admitted to four public university hospitals. The hospitals included were; Jai Prakash Narayan Apex Trauma Center (JPNATC), connected to the All India Institute of Medical Sciences in New Delhi, a large centre soley dedicated to trauma care; King Edward Memorial hospital (KEM) in Mumbai, a tertiary level hospital but without dedicated truma wards; Lokmanya Tilak Municipal General Hospital (LTMGH) in Mumbai, a tertiary lever public university hospital with a smaller dedicated truma ward; and Seth Sukhlal Karnani Memorial Hospital (SSKM) in Kolkata, connected to The Institute of Post-Graduate Medical Education and Research, a tertiary level public university hospital without a ward dedicated soley to trauma.


## Participants
<!-- 
5a) Specify key elements of the study setting (e.g., primary care, secondary care,
general population) including number and location of centres. [D;V]
5b) Describe eligibility criteria for participants. [D;V]
5c) Give details of treatments received, if relevant.  [D;V]
Patients included in the TITCO cohort were those presenting to one of the participating centres with traumatic injury following a transport accident, fall, assault, self‐harm or burn, and who were alive on arrival and admitted to the hospital for treatment. Patients with an isolated limb injury were excluded from the database as such patients are treated by orthopaedic surgeons and not within trauma care pathway, which comprises a first survey done by a surgical resident with an on‐call consultant surgeon and subsequent observation or surgery. Patients from the TITCO cohort who were aged 15 years or older and underwent surgical intervention within 24 h of arrival were included in this study.
-->

The TITCO cohort include patients with a history of trauma who either got admitted to one of the participating hospitals or who died between arrival and admission. Patients with isolated injurys to limbs and that therefore were treated by orthopaedics and not within the general trauma care were excluded from the database alwell as patients who were dead on arrival.


## Outcome
<!--
6a) Clearly define the outcome that is predicted by the prediction model, including how
and when assessed. [D;V]
6b) Report any actions to blind assessment of the outcome to be predicted. [D;V]
-->
The primary outcome of interest was admission to the ICU during hospitalization. __Any secondary outcomes?__

<!-- Regarding blinding, data on the outcome was collected after data on predictors, i.e. the outcome was not known (in the majority of cases at least, when the predictor data was collected. So in that way the data collectors were "blinded" to the outcome during data collection. So no deliberate action was taken or needed to be taken to avoid bias and therefore nothing should be written about it? -->


## Predictors
<!--
7a) Clearly define all predictors used in developing or validating the multivariable
prediction model, including how and when they were measured. [D;V]
7b) Report any actions to blind assessment of predictors for the outcome and other
predictors. [D;V]
-->
For each patient included in the study the qSOFA score was calculated using data recorded on arrival to the hospital. The calculation of the qSOFA score includes a respiratory rate above 22, GCS below 15 and a systolic blood pressure below 100 where one point is awarded for meeting each of the specified criterias and thus yields a score of 0 to 3. 

## Sample size
<!--
8) Explain how the study size was arrived at [D;V]
-->

We included all eligible patients in the TITCO cohort. <!-- Should this be justified in any way that we found this to be a suffient amount of datapoints for our aim? should we later add how many that turned out to be?-->

## Missing data
<!--
9) Describe how missing data were handled (e.g., complete-case analysis, single
imputation, multiple imputation) with details of any imputation method. [D;V]
-->
We will use complete-case analysis. 

<!--
To adress missing data multiple imputation was used. __How specific should this description be? should there be an explanation on how many regressions where used and the size of the random samples yielding the regressions? programs used? Are there any protocols thats usualy used that i can refer to? i know roughly what multiple imputation is but no idea how to implement it...__ 
-->
<!-- Multiple imputation is quite advanced so I suggest we use a complete case analysis, i.e. we exclude patients with missing data-->
<!-- 


## Statistical analysis methods
<!--
10c. For validation, describe how the predictions were calculated. [V]
10d) Specify all measures used to assess model performance and, if relevant, to
compare multiple models [D;V]
10e. Describe any model updating (for example, recalibration) arising from the validation, if done. [V]
-->

We used R for all statistical analysis [RStudio]. We describe the sample characteristics using counts and percentages for qualitative variables and medians and interquartile ranges (IQR) for quantitative variables. The study sample was randomly split into training, validation, and test samples with 60%, 20%, and 20% of the observations in each sample respectively. We used the training sample to update qSOFA by reestimating the coefficients of the original predictors using logistic regression. We used the validation sample to identify optimal cutoffs - those who maximised the Youden index - for the original and updated qSOFA. We used the test sample to assess and compare the performance of the two models. Bootstrapping was used to estimate 95% confidence intervals associated with point estimates. 

<!-- Read other prediction model papers as well as methodological guides to see what domains of predictive performance that we want to look at -->

Från PPF "This will be a retrospective analysis of a cohort of trauma patients admitted to four public university hospitals in urban India between 2013 and 2015. The complete cohort includes 16 000 patients. The primary outcome will be ICU admission. The predictors included in qSOFA are respiratory rate, Glasgow coma scale, and systolic blood pressure. Validity will be assessed in terms of predictive performance, which in turn will be measures as discrimination and calibration. Calibration will further be visualised using calibration plots. An optimal cutoff will be identified using the Youden index and sensitivity, specificity, precision, recall, positive and negative predictive values will assessed at this cutoff. The original model will finally be compared to a updated model, using the same predictive performance measures. Updating will be performed using logistic regression. Missing data will be handled using multiple imputation. "


<!--*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*-->

Results
========
The TITCO cohort included `r results$n.cohort` patients (Figure 1). We excluded `r results$n.younger.than.18` patients because they were under the age of 18  and `r  results$n.incl2 ` patients because they died between arrival and admission. We further excluded  `r results$n.NA_TOT` patients due to missing data in either admission to the ICU (`r results$n.NA_ICU`), systolic blood pressure  (`r results$n.NA_SBP`), respiratory rate (`r results$n.NA_RR`), or GCS (`r results$n.NA_GCS`). The final sample included `r results$n.complete` patients. 

The median age was `r tabOne$ContTable$Overall["Age","mean"]` with predominantly `r tolower(most.common.sex)` participants (`r p.sex`%). The most common type of injury was `r tolower(most.common.mechanism)`s with `r p.mechanism`% followed by `r tolower(second.most.common.mechanism)`s and `r tolower(third.most.common.mechanism)` accounting for
`r p.mechanism.second` and `r p.mechanism.third` respectively. Furthermore `r  p.transported`% of patients were transported to a participating hospital from another health care facility. Overall in the study population `r p.ICU`% of patients had to be treated for some time in the ICU. The data is presented in \@ref(tab:table1) with corresponding standard deviations. 

```{r figure1, fig.align = 'center', out.width = "100%", fig.cap = "Flowchart of exclussion from the study", echo = FALSE}
knitr::include_graphics("figure1.svg")
```
```{r table1, echo = FALSE}
 kableone(CreateTableOne(data = results$data.table1, strata = "Admitted to the ICU", addOverall = TRUE),caption = "Sample characteristics", label = "table1")
```


We identified the following revised optimal cutoffs: `r boot.cis$cut.sbp[["pe"]]` for SBP, `r boot.cis$cut.rr[["pe"]]` for RR and `r boot.cis$cut.gcs[["pe"]]` for GCS. The corresponding confidence intervals and updated odds ratios can be seen in table 2 together with the original OR from Seymour et al. [källa].


|           | Original OR     | Updated OR(95%CI)         | Optimal cutoffs (95%CI) |
|-----------|-----------------|----------------------------|-------------------------|
| Intercept | yy              | `r ors[["(Intercept)"]]`   |  -                      |
| SBP       | 2.61(2.40-2.85) | `r ors[["new.sbp_score"]]` |`r CIs$cut.sbp`          |
| RR        | 3.18(2.89-3.50) | `r ors[["new.rr_score"]]`  |`r CIs$cut.rr`           |
| GCS       | 4.31(9.96-4.69) | `r ors[["new.gcs_score"]]` |`r CIs$cut.gcs`          |
Table: Table 2 Original and updated OR for the logistic regression model 

<!-- there is something wrong with the code... after running 1000 bootstraps i got this result.
ors[["new.rr_score"]]
[1] "0.783 (-830742.207 - 760171.166)"
$updated.coefs.new.rr_score
[1] "-0.244 (-4.269 - 1.981)" 


and when going to the raw data from the result of updated.coefs.new.rr_score the mean(0.6555935) & median(0.4695019) and max(15.83363)/min(-1.120455). and for ors.new.rr_score: mean(35287.09) & median(1.599198) and max(7524136)/min(-0.3261313)


this makes me think that we get some few outliers that skew the results massivley probably when the cut.rr is so high that only one patient qualify making the coefficient high, in 11 cases the coeff was over 10 with a cut.rr [34 32 34 36 40 34 36 34 34 34 36] and only 11 cases where the OR was over 20 and then it was like 10^6... i dont realy understand how the boot.ci() works but somehow the CIs are absolutely out of control when we have these kind of outliers... 

also regarding the ors for sbp_score:
ors[["new.sbp_score"]]
[1] "1.269 (-0.715 - 1.833)"

somehow the CIs gives a negative value which shouldnt be possible and when i look at the raw results from the bootstrap only one out of 1000 have a OR < 1. mean(1.979518),median(1.88884), max(4.191026), min(0.9627057). why is this the case that the calculated CIs are so different.


-->

The estimated probability of ICU admission and corresponding OR given the OR from table 2 and the distribution of the individual parameters in the study sample are presented in table 3 and 4.  

| Updated score   | Estimated probability of ICU admission(95%CI) | Odds ratio(95% CI)          |
|-----------------|-----------------------------------------------|-----------------------------|
| 0               | `r CIs$est.prob.sum.new.none `                | 1                           |
| 1               | `r CIs$est.prob.sum.new.one `                 |`r CIs$est.OR.sum.new.one `  |
| 2               | `r CIs$est.prob.sum.new.two `                 |`r CIs$est.OR.sum.new.two `  |
| 3               | `r CIs$est.prob.sum.new.three `               |`r CIs$est.OR.sum.new.three ` |

Table: Table 3 Estimated probabilities of ICU admission given an updated qSOFA score of 0-3 and corresponding OR.

| Original score | Estimated probability of ICU admission(95%CI) | Odds ratio(95% CI)          |
|-----------------|-----------------------------------------------|-----------------------------|
| 0               | `r CIs$est.prob.sum.org.none `                | 1                           |
| 1               | `r CIs$est.prob.sum.org.one `                 |`r CIs$est.OR.sum.org.one `  |
| 2               | `r CIs$est.prob.sum.org.two `                 |`r CIs$est.OR.sum.org.two `  |
| 3               | `r CIs$est.prob.sum.org.three `               |`r CIs$est.OR.sum.org.three ` |

Table: Table 4 Estimated probabilities of ICU admission given an original qSOFA score of 0-3 and corresponding OR.


Figure 2? shows a comparison between the estimated and observed probabilities for the new and updated model and table 5 compares the performance in terms of Integrated calibration index (ICI) and Area under the curve (AUC) for the receiver operating characteristics for the original and updated model.

Figure: __FIGURE with 4 figures showing the ICI plots for new vs org and sum vs individual. possibly with CIs inluded in the graph__

Figure: __FIGURE with 2 figures showing the AUC plots for new vs org__

<!-- These are the figures i think would give clarity about the data and the new model, is there a good way to do it? I added a part in the end of the main code-document generating one of the plots in question but im not sure if thats the best way or if we could add CIs in some way in the plot.  -->


Table: Table 5. Comparison of performance, ICI and AUC, between the new and original model as well as the ICI for the combined score for the original and updated.
|              | ICI(95% CI)         | Diff ICI (95% CI)          | AUC (95%CI)     | Diff AUC              |
|--------------|---------------------|----------------------------|-----------------|-----------------------|
| ORG separate | `r CIs$ICI.org`     | -                          | `r CIs$auc.org` |                       |
| NEW seperate | `r CIs$ICI.new`     | `r CIs$diff.ici.qsofa`     | `r CIs$auc.new` |`r CIs$diff.auc.qsofa` |
| ORG sum      | `r CIs$ICI.sum.org` | -                          | -               |                       |
| New sum      | `r CIs$ICI.sum.new` | `r CIs$diff.ici.qsofa.sum` | -               |                       |
Abbreviations: AUC Area under the curve, ICI Integrated calibration index


Discussion
========

<!-- aim and how it went, --->
The aim of this study was to asses the validity of the qSOFA score in predicting ICU admissions in trauma patients. In the context of urban Indian hospitals the original qSOFA score with its original cutoffs and coefficients did not accurately predict ICU admission. The original model underestimated the probability of ICU admission as can be seen in _FIGURE showing ICI plots_. When updating the model adjusting the cutoffs and coefficients the new model performs significantly better both regarding the ICI and AUC.


<!-- usage and implication -->
The estimated probabilities for the new model ranges from `r boot.cis$est.prob.sum.new.none[["pe"]]` to `r boot.cis$est.prob.sum.new.three[["pe"]]`. Where the highest value, corresponding to a score of three, can be said to be a clear overestimation and not accounting for a large portion of patients. A more realistic span is that off 0-2 points, or a probability of `r boot.cis$est.prob.sum.new.none[["pe"]]` to `r boot.cis$est.prob.sum.new.two[["pe"]]` percent. Because of this relatively high base probability of `r boot.cis$est.prob.sum.new.none[["pe"]]`% the updated model can not in itself be used to exclude patients with a low score from the future need of intensive care. Nor can it with certainty predict that patients with a high score will be admitted to the ICU although there exists a significantly higher risk that they will.  

Contrast this to the original use of the qSOFA score where a score of 0 correspond to a mortality of 1  <!-- approx taken from figure --> percent and a score of 2-3 correspond to a much higher mortality of 7-20 %<!-- taken from figure ish-->. This makes the original qSOFA in the context of screening for risk of dying due to severe infections a good tool to sort patients with a low score to also be of a low risk of dying.

<!-- It can also be noted that out of the three parameters used in the updated model only GCS was significantly associated with a higher risk of ICU admission. This can be partially explained by that new cutoffs where calculated for each bootstrap resulting in different cutoffs being evaluated resulting in an increased uncertainty. It could also be that RR and SBP does not have a significant correlation to ICU admission. Although the data suggest that there is a correlation and therefore a possibility of using it in the formation of a predictive model the result is not significant. --> <!-- körde om bootstrapen igen och då fick jag ett annat resultat så stryker denna för nu... -->

<!--limitations --->

Because of the composition and context of the TITCO database the results are likely to be generalizable to similar settings, that is to other urban hospitals of similar economic and resource level. <!-- Why do you think that?,,, similar settings should result in similar results? or should i formulate it in another way?-->. It is not clear whether the results can be used in rural settings or setting with higher or lower resource levels. 

It should also be noted that `r  p.transported` of the patients where for an unspecified time first treated at another hospital and then later transferred to one of the participating hospitals. Because of this the parameters used in this study only shows the first ones taken at arrival to the participating hospital and not always the first ones taken after the injury and does not take into account any interventions performed before arrival or between arrival and admission to the ICU.

Furthermore out of the `r results$n.adults` patients above the age of 18 only `r results$n.complete` had complete data, or `r results$n.complete/results$n.adults`%. With this amount of missing data it is unclear if the data can be said to be missing completely at random or if there are any underlying factors contributing to the missing of data that could skew the results leading to a higher uncertainty regardless of bootstraping<!-- Does the amount of missing data really say anything about the pattern of missing data?--> <!-- No and thas what i want to caspulate that we can not now... so I still want to add something about the amount of missing data but not sure how... --> .

[@Miyamoto2020] Comparing this to the results by Miyamoto on trauma patients in Japan where they looked at the qSOFA scores predictive power on mortality they showed that in their sample an elevated qSOFA score of 2-3 upon arrival to the hospital had an odds ratio compared with that of a score of 0 of 2.92-4.84 depending on the prehospital score. They also calculated that in their sample qSOFA had an AUC of 0.69-0.76.

Another comparrison can be made to that of the development In its original context the qSOFA model had an AUC for mortality of 0.66 for patients treated at the ICU and 0.81 for patients without ICU encounters. The original OR for the parameters can be seen in table 2. (orginalkällan) 

Although these studies looked at mortality and not ICU admission one can see that both the mentioned studies had a comparable but slightly higher AUC and a higher OR than can be shown in this study. This is to be somewhat expected since it can be argued that although there is a connection between ICU admission and mortality through severity of that patients condition there were patients who died without being treated in the ICU. The reason for this is not certain but one can argue that this is because the patients condition was such that the patients either died before they could be admitted to the ICU, the doctors decided that intensive care would not be of benefit for the patient, or due to a lack of resources or ICU beds such that patients had to be prioritized for intensive care and not all patients who would benefit from it got admitted to the ICU. For context the number of ICU beds per 100 000 people is 2.3 in India compared with for example 7.3 in Japan. (Källa 19)

<!--It's a bit all over the place right now lacking a common thread but i wanted to add all the parts first and then adjust and rearange depending on your feedback --> 


## Questions:

From coordinator:

* Why are hospitals without trauma wards included?(in the TITCO cohort)

Most hospitals in India, like in Sweden, don't have trauma specific wards. They still treat large volumes of trauma patients.

* Some more consideration of risks would be good [in ethical consideration] -- especially the potential to stigmatize vulnerable participants [i.e. low socioeconomic status]. How will this be avoided?

Not sure about this. Can you reason about what stigma that might be?

Mine: 
* Baseline risk and coefficient corresponding to that. Is it really best to take it directly from the original paper as a fair comparison or should it be adjusted for our data to get a better perspective for discussion? And I only found a figure in the supplementary material and no numbers, should i estimate what can be read from the figure or is there a better way, it was however really small just a few percent?

I would estimate it from the figure. The idea would be to see if the original score works, and therefore no adjustments are necessary, or if one needs to use an updated score, like the one you're developing. Only updating the baseline risk is somewhere in between these two options, which is also possible, but would add a third model.

