---
title: The Title of Your Study Here
subtitle: Add a Subtitle if Needed
author: Björn Lindberg
bibliography: bibliography.bib
csl: bmcemerg.csl
output: pdf_document
---


Introduction 
============
Trauma is defined as the clinical entity composed of both physical injury and the body’s associated response. [MeSH] Trauma patients are a heterogeneous group with a wide range of different causes the most prominent ones are traffic related injuries, falls, self-harm, interpersonal violence and burns. Trauma is a global health problem and the total number of patients with injuries is estimated to be at least 520 million per year. [@James2020] 

The effects of trauma on patients and their families can be divided into non-fatal and fatal trauma. The effects of non-fatal trauma can be subdivided into Disability Adjusted Life Years (DALY) lost and economical effects. DALY is a measurement using both the mortality in years lost between mean age and death and morbidity as in years lived with reduced function due to the disease. In 2017 trauma accounted for 13% of total DALY lost globally. [@GBD2017DALY] The economic effects are both out of pocket expenses for treatment as well as reduced ability to work after injury. A study in northern India from 2019 calculated that 22% of patients who presented with traumatic injuries afterward had 30% or more of their total expenditure on healthcare related costs and 12 % were pushed below the international poverty line as a consequence. [@Prinja2019]

Regarding the effect of trauma on mortality in 2017 trauma accounted for 8% of total global deaths. [@GBD2017DALY] According to calculations done by Mock et al. [@Mock2012] 90% of all deaths due to trauma were in low and middle income countries which also had a fatality rate 60-80% higher than that of high income countries and an improvement in trauma care globally to the level of that in high income countries could save about 1.7 lives annually, or about one third of the global deaths due to trauma.

It is known that most fatalities in response to traumatic injuries occur shortly after the injury and to reduce the mortality in patients with severe trauma requires early intervention, often surgical, as by the “golden hour” concept. This is often due to patients suffering from internal and external bleeding where surgery is an important part of the hemorrhage control. [@King2019] In parts of the world with fewer hospitals, reduced access to health care and less developed prehospital systems the time parameter is one of the contributing factors to higher mortality.

For patients with severe infections the Sequential Organ Failure Assessment (SOFA) score was developed in 1994 to grade the level of organ dysfunction in critically ill patients. The SOFA score is based on the reduced function scored 0-4 for each of the respiratory, cardiovascular, neurological, hepatic, coagulation and renal systems. Initially the focus was mainly patients suffering from severe infections and sepsis, but it was early recognized to be applicable in a wide range of critical illnesses. [@Lambden2019] 

Alongside the last revision of the sepsis definition in 2016 the quick Sequential Organ Failure Assessment score (qSOFA score) was developed. The qSOFA score includes three parameters: Systolic blood pressure (SBP) below 100 mmHg, a respiratory rate (RR) above 22 min and an altered consciousness, measured with a Glasgow Coma Scale (GCS) score below 15. One point is awarded for each parameter yielding a total score of 0-3. Introduced as a screening tool, to be performed bedside to identify patients with high-risk of organ dysfunction due to infections. [@David2017;@Shankar2016]qSOFA has lately also been shown to predict mortality in trauma patients both in prehospital and hospital settings [@Miyamoto2019; @Miyamoto2020].

Another tool to assess the severity of a patient’s condition due to trauma is the Revised Trauma Score (RTS) and the Triage Revised Trauma Score (T-RTS), which include the same parameters as qSOFA - GCS, SBP and RR - but with a score of 0-4 for each parameter instead of a single criteria. The scores can either be presented individually as used in RTS or added together to a total of 0-12 used in the T-RTS. RTS was intended for outcome evaluation and assessment of injury severity and the T-RTS for triage and finding patients a prehospital or initial setting in most need of specialized trauma care [@Champion1989] However there are some articles raising concerns about the sensitivity of T-RTS depending on the type of injury. [@Giannakopoulos2011;@Alvarez2016]

Because of the inherent urgency and complexity in trauma care it is vital to have an initial scoring system that is both simple to perform and that does not need laboratory nor other advanced measuring equipment. For this reason, the SOFA score can be said to use too advanced techniques and the T-RTS with its many categories to be too complicated. The choice of using qSOFA is thus because of its promising signs in other studies and because of the possible clinical advantage in its simplicity.  

It should also be noted that the majority of the research done on the topic of trauma and trauma care has been performed in high income countries. It is thus not clear if the results are also applicable in low resource settings, which arguably stand to benefit the most of a scoring system to be able to prioritize the limited resources at hand. Furthermore, most studies report the predictive power of different scoring systems on mortality and it has not yet been sufficiently studied whether an initial qSOFA score can be used to predict admissions to an intensive care unit and thus aid in the direct clinical treatment of trauma patients.

<!-- just for the KI paper ---> 
## Healthcare in India 

The Indian healthcare is of now severely underfunded, the national healthcare expenditure is about 3.5% of national GDP, or about 70 USD per person in 2017, compared to a national healthcare expenditure of 5900 USD in Sweden for the same year. Out of these 70 USD per person 27 % is government expenditure and 72% is private expenditure, mostly out of pocket expenditures. (https://data.worldbank.org/) This makes the unregulated private healthcare sector almost 3 times as large as the public sector providing 80% of outpatient care and 60% of inpatient care in India. India also lacks healthcare personal and in many cases subspecialized medical training, especially regarding trauma care. (Nobhojit Roy)(Martin Gerdin)

Besides the deficient monetary aspects of Indias healthcare system it can also be said that there is a lack of sufficiently enforced safety policies and regulations, for example road traffic and workplace safety. Although it should be noted that in the new national health policy from 2017 one of the primary focuses are preventing rail and road traffic accidents. The national health policy also aims at expanding the number of hospital beds dedicated to trauma care across the country reducing the time between injury and care as guided by the golden hour concept discussed above. (Nobhojit Roy)(Martin Gerdin) 


## Aim  
The aim of this project is to assess the validity of qSOFA in predicting ICU admission in trauma patients admitted to hospitals in a low resource setting using data from the TITCO cohort [TITCO].


Methods
=======

## Source of data

<!-- 
4a) Describe the study design or source of data (e.g., randomized trial, cohort, or
registry data), separately for the development and validation data sets, if applicable
4b) Specify the key study dates, including start of accrual; end of accrual; and, if
applicable, end of follow-up. 

This study involved an analysis of the observational cohort Towards Improved Trauma Care Outcomes in India (TITCO), for which data were collected before the conception of this study. Ethics committees at all participating centres approved the collation of the database and granted a waiver of consent for patients with trauma (Lokmanya Tilak Municipal General Hospital, IEC/11/13; King Edward Memorial Hospital, IEC(I)/OUT/222/14; Seth Sukhlal Karnani Memorial Hospital, IEC/279; All‐India Institute of Medical Sciences, IEC/NP‐279/2013 RP‐01/2013). The study was conducted using anonymized data, and was registered at ClinicalTrials.gov (NCT03069755) before the research was undertaken.

-->

We conducted a retrospective analysis of the observational Towards Improved Trauma Care Outcomes in India (TITCO) cohort [@TITCO] <!-- TITCO collaborators (2017). TITCO dataset version 1. Available from https://github.com/titco/titco-I. (Behöver läggas in i .bib)-->. The data for TITCO was collected during july 2013 to december 2015 and contains patients admitted to four public university hospitals. The hospitals included were; Jai Prakash Narayan Apex Trauma Center (JPNATC), connected to the All India Institute of Medical Sciences in New Delhi, a large centre soley dedicated to trauma care; King Edward Memorial hospital (KEM) in Mumbai, a tertiary level hospital but without dedicated truma wards; Lokmanya Tilak Municipal General Hospital (LTMGH) in Mumbai, a tertiary lever public university hospital with a smaller dedicated truma ward; and Seth Sukhlal Karnani Memorial Hospital (SSKM) in Kolkata, connected to The Institute of Post-Graduate Medical Education and Research, a tertiary level public university hospital without a ward dedicated solely to trauma.


## Participants
<!-- 
5a) Specify key elements of the study setting (e.g., primary care, secondary care,
general population) including number and location of centres. [D;V]
5b) Describe eligibility criteria for participants. [D;V]
5c) Give details of treatments received, if relevant.  [D;V]
Patients included in the TITCO cohort were those presenting to one of the participating centres with traumatic injury following a transport accident, fall, assault, self‐harm or burn, and who were alive on arrival and admitted to the hospital for treatment. Patients with an isolated limb injury were excluded from the database as such patients are treated by orthopaedic surgeons and not within trauma care pathway, which comprises a first survey done by a surgical resident with an on‐call consultant surgeon and subsequent observation or surgery. Patients from the TITCO cohort who were aged 15 years or older and underwent surgical intervention within 24 h of arrival were included in this study.
-->

The TITCO cohort include patients with a history of trauma who either got admitted to one of the participating hospitals or who died between arrival and admission. Patients with isolated injurys to limbs and that therefore were treated by orthopaedics and not within the general trauma care were excluded from the database aswell as patients who were dead on arrival. 


## Outcome
<!--
6a) Clearly define the outcome that is predicted by the prediction model, including how
and when assessed. [D;V]
6b) Report any actions to blind assessment of the outcome to be predicted. [D;V]
-->
The primary outcome of interest was admission to the ICU during hospitalization. __Any secondary outcomes?__

<!-- Regarding blinding, data on the outcome was collected after data on predictors, i.e. the outcome was not known (in the majority of cases at least, when the predictor data was collected. So in that way the data collectors were "blinded" to the outcome during data collection. So no deliberate action was taken or needed to be taken to avoid bias and therefore nothing should be written about it? -->


## Predictors
<!--
7a) Clearly define all predictors used in developing or validating the multivariable
prediction model, including how and when they were measured. [D;V]
7b) Report any actions to blind assessment of predictors for the outcome and other
predictors. [D;V]
-->
For each patient included in the study the qSOFA score was calculated using data recorded on arrival to the hospital. The calculation of the qSOFA score includes a respiratory rate above 22, GCS below 15 and a systolic blood pressure below 100 where one point is awarded for meeting each of the specified criterias and thus yields a score of 0 to 3. 

## Sample size
<!--
8) Explain how the study size was arrived at [D;V]
-->

We included all patients in the TITCO cohort above the age of 18 who was alive at admission to the hospital except patients with externally controlled breathing upon arrival, either through intubation or a surgical airway. <!-- Is this the correct place to add exclussion critera or should it be up on participants instead??-->

## Missing data
<!--
9) Describe how missing data were handled (e.g., complete-case analysis, single
imputation, multiple imputation) with details of any imputation method. [D;V]
-->
We will use complete-case analysis. 


## Statistical analysis methods
<!--
10c. For validation, describe how the predictions were calculated. [V]
10d) Specify all measures used to assess model performance and, if relevant, to
compare multiple models [D;V]
10e. Describe any model updating (for example, recalibration) arising from the validation, if done. [V]
-->

We used R for all statistical analysis [RStudio]. We describe the sample characteristics using counts and percentages for qualitative variables and medians and interquartile ranges (IQR) for quantitative variables. The study sample was randomly split into training, validation, and test samples with 60%, 20%, and 20% of the observations in each sample respectively. We used the training sample to update qSOFA by reestimating the coefficients of the original predictors using logistic regression. We used the validation sample to identify optimal cutoffs - those who maximised the Youden index - for the original and updated qSOFA. We used the test sample to assess and compare the performance of the two models. Bootstrapping was used to estimate 95% confidence intervals associated with point estimates. 

<!-- Read other prediction model papers as well as methodological guides to see what domains of predictive performance that we want to look at -->

Från PPF "This will be a retrospective analysis of a cohort of trauma patients admitted to four public university hospitals in urban India between 2013 and 2015. The complete cohort includes 16 000 patients. The primary outcome will be ICU admission. The predictors included in qSOFA are respiratory rate, Glasgow coma scale, and systolic blood pressure. Validity will be assessed in terms of predictive performance, which in turn will be measures as discrimination and calibration. Calibration will further be visualised using calibration plots. An optimal cutoff will be identified using the Youden index and sensitivity, specificity, precision, recall, positive and negative predictive values will assessed at this cutoff. The original model will finally be compared to a updated model, using the same predictive performance measures. Updating will be performed using logistic regression. Missing data will be handled using multiple imputation. "


<!--*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*-->

Results
========
The TITCO cohort included `r results$n.cohort` patients (Figure 1). We excluded `r results$n.younger.than.18` patients because they were under the age of 18  and `r  results$n.incl2 ` patients because they died between arrival and admission. We further excluded  `r results$n.NA_TOT` patients due to missing data in either admission to the ICU (`r results$n.NA_ICU`), systolic blood pressure  (`r results$n.NA_SBP`), respiratory rate (`r results$n.NA_RR`), or GCS (`r results$n.NA_GCS`). The final sample included `r results$n.complete` patients. 

The median age was `r tabOne$ContTable$Overall["Age","mean"]` with predominantly `r tolower(most.common.sex)` participants (`r p.sex`%). The most common type of injury was `r tolower(most.common.mechanism)`s with `r p.mechanism`% followed by `r tolower(second.most.common.mechanism)`s and `r tolower(third.most.common.mechanism)` accounting for
`r p.mechanism.second` and `r p.mechanism.third` respectively. Furthermore `r  p.transported`% of patients were transported to a participating hospital from another health care facility. Overall in the study population `r p.ICU`% of patients had to be treated for some time in the ICU. The data is presented in \@ref(tab:table1) with corresponding standard deviations. 

```{r figure1, fig.align = 'center', out.width = "100%", fig.cap = "Flowchart of exclussion from the study", echo = FALSE}
knitr::include_graphics("figure1.svg")
```
```{r table1, echo = FALSE}
 kableone(CreateTableOne(data = results$data.table1, strata = "Admitted to the ICU", addOverall = TRUE),caption = "Sample characteristics", label = "table1")
```


We identified the following revised optimal cutoffs: `r boot.cis$cut.sbp[["pe"]]` for SBP, `r boot.cis$cut.rr[["pe"]]` for RR and `r boot.cis$cut.gcs[["pe"]]` for GCS. The corresponding confidence intervals and updated odds ratios can be seen in table 2 together with the original OR from Seymour et al. [källa].


|           | Original OR     | Updated OR(95%CI)         | Optimal cutoffs (95%CI) |
|-----------|-----------------|----------------------------|-------------------------|
| Intercept | yy              | `r ors[["(Intercept)"]]`   |  -                      |
| SBP       | 2.61(2.40-2.85) | `r ors[["new.sbp_score"]]` |`r CIs$cut.sbp`          |
| RR        | 3.18(2.89-3.50) | `r ors[["new.rr_score"]]`  |`r CIs$cut.rr`           |
| GCS       | 4.31(9.96-4.69) | `r ors[["new.gcs_score"]]` |`r CIs$cut.gcs`          |
Table: Table 2 Original and updated OR for the logistic regression model 


The estimated probability of ICU admission and corresponding OR given the OR from table 2 and the distribution of the individual parameters in the study sample are presented in table 3 and 4.  

| Updated score   | Estimated probability of ICU admission(95%CI) | Odds ratio(95% CI)          |
|-----------------|-----------------------------------------------|-----------------------------|
| 0               | `r CIs$est.prob.sum.new.none `                | 1                           |
| 1               | `r CIs$est.prob.sum.new.one `                 |`r CIs$est.OR.sum.new.one `  |
| 2               | `r CIs$est.prob.sum.new.two `                 |`r CIs$est.OR.sum.new.two `  |
| 3               | `r CIs$est.prob.sum.new.three `               |`r CIs$est.OR.sum.new.three ` |

Table: Table 3 Estimated probabilities of ICU admission given an updated qSOFA score of 0-3 and corresponding OR.

| Original score | Estimated probability of ICU admission(95%CI) | Odds ratio(95% CI)          |
|-----------------|-----------------------------------------------|-----------------------------|
| 0               | `r CIs$est.prob.sum.org.none `                | 1                           |
| 1               | `r CIs$est.prob.sum.org.one `                 |`r CIs$est.OR.sum.org.one `  |
| 2               | `r CIs$est.prob.sum.org.two `                 |`r CIs$est.OR.sum.org.two `  |
| 3               | `r CIs$est.prob.sum.org.three `               |`r CIs$est.OR.sum.org.three ` |

Table: Table 4 Estimated probabilities of ICU admission given an original qSOFA score of 0-3 and corresponding OR.


Figure 2? shows a comparison between the estimated and observed probabilities for the new and updated model and table 5 compares the performance in terms of Integrated calibration index (ICI) and Area under the curve (AUC) for the receiver operating characteristics for the original and updated model.

Figure: __FIGURE with 4 figures showing the ICI plots for new vs org and sum vs individual. possibly with CIs inluded in the graph__

Figure: __FIGURE with 2 figures showing the AUC plots for new vs org__

<!-- These are the figures i think would give clarity about the data and the new model, is there a good way to do it? I added a part in the end of the main code-document generating one of the plots in question but im not sure if thats the best way or if we could add CIs in some way in the plot.  -->


Table: Table 5. Comparison of performance, ICI and AUC, between the new and original model as well as the ICI for the combined score for the original and updated.
|              | ICI(95% CI)         | Diff ICI (95% CI)          | AUC (95%CI)     | Diff AUC              |
|--------------|---------------------|----------------------------|-----------------|-----------------------|
| ORG separate | `r CIs$ICI.org`     | -                          | `r CIs$auc.org` |                       |
| NEW seperate | `r CIs$ICI.new`     | `r CIs$diff.ici.qsofa`     | `r CIs$auc.new` |`r CIs$diff.auc.qsofa` |
| ORG sum      | `r CIs$ICI.sum.org` | -                          | -               |                       |
| New sum      | `r CIs$ICI.sum.new` | `r CIs$diff.ici.qsofa.sum` | -               |                       |
Abbreviations: AUC Area under the curve, ICI Integrated calibration index


Discussion
========

<!-- aim and how it went, --->
The aim of this study was to asses the validity of the qSOFA score in predicting ICU admissions in trauma patients. We found that in the context of urban Indian hospitals the original qSOFA score with its original cutoffs and coefficients did not accurately predict ICU admission. The original model underestimated the probability of ICU admission as can be seen in _FIGURE showing ICI plots_. When updating the model adjusting the cutoffs and coefficients the new model performs significantly better in terms of ICI and slightly better in terms of AUC. 


<!-- Strenghts -->
  

<!-- usage and limitation -->


While developing the original qSOFA score as a screening tool for patients with an increased risk of dying from severe infection it was shown that a low score of 0-1 corresponded to a mortality of approximately 1-2 percent and that a high score of 2-3 corresponded to a significantly higher mortality of 7-20 %. When using the predetermined cutoff value of 2 points this result made the original qSOFA a useful tool to be able to sort patients with a low score to also be of a low risk of dying. On the other hand the results also allowed for that patients with a higher score could be more closely monitored and get access to treatment before more severe symptoms developed. 

Contrast this to the results from the updated model used in the context of predicting ICU admission of trauma patients as have been done in this study. Our findings shows that a low score of 0-1 correspond to a probability for ICU admission of `r boot.cis$est.prob.sum.new.none[["pe"]]`-`r boot.cis$est.prob.sum.new.one[["pe"]]`% and a high score of 2-3 correspond to a mortality of `r boot.cis$est.prob.sum.new.two[["pe"]]`-`r boot.cis$est.prob.sum.new.three[["pe"]]` %. This shows that regardless of the score the patients still have a relatively high risk for ICU admission. Although a higher score is shown to correlate with a higher risk of ICU admission it is not clear what clinical relevance this score would have in the process of deciding the adequate level of care for patients with severe injuries. <!-- I think this better helps me to get my point across that the results are not clear enough that so it will not be possible to use the updated score in this context the same way as the original score was used the that context. -->

The results also showed that out of the parameters included in the qSOFA respiratory rate was not significantly associated with a higher risk of ICU admission, OR of `r ors[["new.rr_score"]]`. This can be partially explained by that a new cutoff was calculated for each bootstrap resulting in different cutoffs being used while evaluating the model in turn leading to an increased uncertainty. Further research is needed before the original or an updated qSOFA score as presented in this study can be used in clinical desicionmaking.   

Although no other studies have been done correlating qSOFA to ICU admission studies have explored the conection between qSOFA and mortality both regarding patients with risk for sepsis [Orgianlkälla] and trauma patients.[@Miyamoto2020] Both these studies show that there is a clear connection between the qSOFA score and mortality and that the ORs as well as the AUCs is higher when looking at mortality in both studies compared to the results for this study. Although the AUCs only differ slightly. It can be argued that both ICU admission and mortality results from a more severe patient condition and thus will be correlated. However it should be noted that there exists patients who died without being admitted to the ICU reducing this correlation. This could be partly because patient where so severely injured that they either died before being able to be transferred to the ICU or it was decided that intensive care would not be of use for the patients. But it could also be due to a lack of ICU beds in the participating hospitals such that patients had to be prioritized for intensive care and not all patients who would benefit from it got admitted to the ICU. For context the number of ICU beds per 100 000 people is 2.3 in India compared with for example 7.3 in Japan. (Källa 19) To be able to explore this question more thoroughly further research is needed. 

The study had several limitations. First the data collected for the TITCO database came from a homogeneous group of participating hospitals all in urban settings in India. Because of this it is not clear that the results presented here are generalizable to for example hospitals with a rural location or a different economic level. 

Second it should be noted that `r  p.transported` percent of the patients where for an unspecified time first treated at another hospital and then later transferred to one of the participating hospitals. Because of this the parameters used in this study only shows the first ones taken at arrival to the participating hospital and not always the first ones taken after the injury. furthermore the results in this study does not take into account any interventions performed before arrival or between arrival and admission to the ICU.


<!-- future research -->
A way for future research to investigate the usage of qSOFA in ICU admission would be to look at ICU admission, qSOFA score and mortality and investigate whether patients depending on their qSOFA score who where admitted to the ICU had a lower morality compared to patients with the same qSOFA score not admitted to the ICU. The aim of this would be to decide which patients would be the most helped by intensive care to better allocate the resources on hand. 


##Conclusion
An updated qSOFA score can be used to estimate the overall risk for ICU admission in trauma patients in urban India hospitals but more research is needed before it can be implemented and used in clinical decision making.



<!-- Add strengths and limitations, implications for future research and practice, and conclusions -->
